---
title: "plyr and dplyr compatibility"
author: "Kirill Müller"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F}
library(magrittr)
knit_print.data.frame = function(x, ...) {
  res = paste(c('', '', knitr::kable(x)), collapse = '\n')
  knitr::asis_output(res)
}
`knit_print.{` = function(x, ...) {
  res <- paste(c("```r", paste0(c("", rep("  ", length(x)-1)), as.character(x)), "}", "```", ""), collapse = '\n')
  knitr::asis_output(res)
}
```

# Problem statement

When loading both `plyr` and `dplyr`,
the last package loaded overwrites symbols exported by the
package loaded first:

```{r}
library(plyr)
library(dplyr)
```

Currently, the following symbols are affected:

```{r}
plyr_exports <- ls("package:plyr")
dplyr_exports <- ls("package:dplyr")
(both_exports <- intersect(plyr_exports, dplyr_exports))
```

This means that existing projects that use `plyr`
cannot simply load `dplyr` using `library(dplyr)`
without potentially breaking existing code.
There are workarounds, but all of them seem to have
specific disadvantages:

- Load only one of both packages, access functionality from the
  other package via explicit qualification (e.g., `dplyr::summarise`).
    - Cumbersome
- Don't load, always access via explicit qualification
    - Requires extensive code rewriting for existing projects
- Load `dplyr` after `plyr`, modify usage of conflicting symbols
  in the code
    - Doesn't safeguard new code that attempts to use `dplyr` primitives with `plyr`

This document explores alternative solutions.


# Analysis

Let's take a closer look at the interface of the functions exported from both packages:

```{r echo = FALSE}
formals_df <-
  both_exports %>%
  setNames(nm = .) %>%
  lapply(
    . %>% {
      lapply(
        c("plyr", "dplyr") %>% setNames(nm = .),
        function (x) {
          get(., sprintf("package:%s", x)) %>%
            formals %>% names %>% paste(collapse = ", ")
        }
      )
    }
  ) %>%
  plyr::ldply(data.frame, stringsAsFactors = FALSE, .id = "name") %>%
  plyr::mutate(data_first = grepl("^[.]data,", dplyr) | grepl("^df,", plyr),
               identical = plyr == dplyr) %>%
  plyr::arrange(!data_first, !identical, name)

formals_df
```

We can split the overlaps in two groups:

- The first argument is a data frame or table object
- No data frame or table object, but identical formals

## Data is first argument

```{r echo = FALSE}
formals_df %>%
  subset(data_first) %>%
  plyr::mutate(data_first = NULL)
```

Here, `mutate`, `summari[sz]e` and `arrange` mean the same in both `plyr` and `dplyr`,
although `plyr::summari[sz]e` on a grouped `tbl_df` seems to
destroy the grouping.
The `count` function is similar, too, only that `plyr::count`
is more similar to `dplyr::count_`, and that for some reason
the first argument of `dplyr::count` is called `x` and not `.data`.
Also, `plyr::rename` seems to be more similar to `dplyr::rename_`.

## Identical formals

```{r echo = FALSE}
formals_df %>%
  subset(!data_first & identical) %>%
  plyr::mutate(data_first = NULL, identical = NULL)
```

Here, we take a look at the bodies.
We compare them for textual identity.

```{r echo = FALSE}
body_l <-
  formals_df %>%
  subset(!data_first & identical) %>%
  extract2("name") %>%
  as.character %>%
  setNames(nm = .) %>%
  lapply(
    . %>% {
      lapply(
        c("plyr", "dplyr") %>% setNames(nm = .),
        function (x) {
          get(., sprintf("package:%s", x)) %>%
            body
        }
      )
    }
  )
```

```{r}
body_l %>%
  llply(. %>% { identical(as.character(.$plyr), as.character(.$dplyr)) } ) %>%
  ldply(. %>% data.frame(identical_body = .), .id = "name")
```

Only `failwith` is different, but the effect seems to be the same:

```{r}
body_l$failwith$plyr
body_l$failwith$dplyr
```


# Concept

For fully seamless transition between `plyr` and `dplyr`,
a compatibility package looks like a possible option.
It would provide the union of the exported functions of both packages,
and compatibility wrappers for the two functions
`count` and `rename`
that need special attention.

Deprecating the functions `count` and `rename` in the `plyr` package seems simpler.
For an even more radical solution,
all overlapping functions could be deprecated,
referring to identical functionality in `dplyr`:

- `mutate` -> `dplyr::mutate`
    - row names are lost
- `summari[sz]e` -> `dplyr::summari[sz]e`
- `arrange(df, ...)` -> `dplyr::arrange(.data, ...)`
- `count(df, vars, wt_var)` -> `dplyr::count_(x, vars, wt, sort = FALSE)`
    - modified call
    - rename `n` column to `freq`
- `rename(x, replace, warn_missing)` -> `dplyr::rename_(.data, replace)`
    - `replace` needs to be massaged
    - `warn_missing` is always `TRUE`
- `desc()`, `failwith()`, `id()` -> `dplyr::...`


# Summary

For practical use, a thin compatibility layer seems to work reasonably well
for a project that was created using `plyr` and is now transitioning
towards `dplyr`:

```{r echo = FALSE}
setHook(packageEvent("plyr", "attach"), function(...) {}, action = "replace")
devtools::unload(devtools::inst("dplyr"))
devtools::unload(devtools::inst("plyr"))
```

```{r}
library(pdlyr)
```

Tests in the `pdlyr` package will assure that this package
can be safely loaded with `warn.conflicts = FALSE`.

A few usage examples:

```{r}
mtcars %>% mutate(lphkm = 100 * 3.785411784 / 1.609344 / mpg) %>% head
mtcars %>% plyr::mutate(lphkm = 100 * 3.785411784 / 1.609344 / mpg) %>% head
mtcars %>% ddply("cyl", summarize, mean_mpg = mean(mpg))
mtcars %>% ddply("cyl", plyr::summarise, mean_mpg = mean(mpg))
mtcars %>% arrange(-wt) %>% head
mtcars %>% plyr::arrange(-wt) %>% head
mtcars %>% count("gear")
mtcars %>% plyr::count("gear")
mtcars %>% rename(list(mpg = "miles_per_gallon")) %>% extract(1:2) %>% head
mtcars %>% plyr::rename(list(mpg = "miles_per_gallon")) %>% extract(1:2) %>% head
```

The original `plyr` implementations can be accessed via shortcuts:

```{r}
pmutate %>% identical(plyr::mutate)
pcount %>% identical(plyr::count)
prename %>% identical(plyr::rename)
```
