---
title: "Vignette Title"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=F}
library(magrittr)
knit_print.data.frame = function(x, ...) {
  res = paste(c('', '', knitr::kable(x)), collapse = '\n')
  knitr::asis_output(res)
}
```

# Problem statement

When loading both `plyr` and `dplyr`,
the last package loaded overwrites symbols exported by the
package loaded first:

```{r}
library(plyr)
library(dplyr)
```

Currently, the following symbols are affected:

```{r}
plyr_exports <- ls("package:plyr")
dplyr_exports <- ls("package:dplyr")
(both_exports <- intersect(plyr_exports, dplyr_exports))
```

This means that existing projects that use `plyr`
cannot simply load `dplyr` using `library(dplyr)`
without potentially breaking existing code.
There are workarounds, but all of them seem to have
specific disadvantages:

- Load only one of both packages, access functionality from the
  other package via explicit qualification (e.g., `dplyr::summarise`).
    - Cumbersome
- Don't load, always access via explicit qualification
    - Requires extensive code rewriting for existing projects
- Load `dplyr` after `plyr`, modify usage of conflicting symbols
  in the code
    - Doesn't safeguard new code that attempts to use `dplyr` primitives with `plyr`

This document describes a compatibility layer as an alternative solution.


# Analysis

Let's take a closer look at the interface of the functions exported from both packages:

```{r echo = FALSE}
formals_df <-
  both_exports %>%
  setNames(nm = .) %>%
  lapply(
    . %>% {
      lapply(
        c("plyr", "dplyr") %>% setNames(nm = .),
        function (x) {
          get(., sprintf("package:%s", x)) %>%
            formals %>% names %>% paste(collapse = ", ")
        }
      )
    }
  ) %>%
  plyr::ldply(data.frame, stringsAsFactors = FALSE, .id = "name") %>%
  plyr::mutate(data_first = grepl("^[.]data,", dplyr) | grepl("^df,", plyr),
               identical = plyr == dplyr) %>%
  plyr::arrange(!data_first, !identical, name)

formals_df
```

We can split the overlaps in two groups:

- The first argument is a data frame or table object
- No data frame or table object, but identical formals

## Data is first argument

```{r echo = FALSE}
formals_df %>%
  subset(data_first) %>%
  plyr::mutate(data_first = NULL)
```

Here, `mutate`, `summari[sz]e` and `arrange` mean the same in both `plyr` and `dplyr`,
although `plyr::summari[sz]e` on a grouped `tbl_df` seems to
destroy the grouping.
The `count` function is similar, too, only that `plyr::count`
is more similar to `dplyr::count_`, and that for some reason
the first argument of `dplyr::count` is called `x` and not `.data`.
Also, `plyr::rename` seems to be more similar to `dplyr::rename_`.

## Identical formals

```{r echo = FALSE}
formals_df %>%
  subset(!data_first & identical) %>%
  plyr::mutate(data_first = NULL, identical = NULL)
```

Here, we take a look at the bodies.
We compare them for textual identity.

```{r echo = FALSE}
body_l <-
  formals_df %>%
  subset(!data_first & identical) %>%
  extract2("name") %>%
  as.character %>%
  setNames(nm = .) %>%
  lapply(
    . %>% {
      lapply(
        c("plyr", "dplyr") %>% setNames(nm = .),
        function (x) {
          get(., sprintf("package:%s", x)) %>%
            body
        }
      )
    }
  )
```

```{r}
body_l %>%
  llply(. %>% { identical(as.character(.$plyr), as.character(.$dplyr)) } ) %>%
  ldply(. %>% data.frame(identical_body = .), .id = "name")
```

